```{r setup-section-text, echo=FALSE, include=FALSE}
library(tidyverse)
bikes <- readr::read_csv("https://www.cedricscherer.com/data/london-bikes.csv", col_types = "Dcfffilllddddfc")
```

# Working with Text {#text}

## Titles

By default, **ggplot2** uses the column names specified for the aesthetic mappings as titles. That means, one could overwrite the column names in the source data. Usually the names are optimized for developing code (i.e. they contain no white spaces or unusual symbols) and it is better to leave them as they are. Also, the columns might be addressed somewhere else in the script and overwriting their names sounds like a bad idea in such a case.

```{r axis-labs-default}
ggplot(bikes, aes(y = count, x = temp, color = day_night)) + 
  geom_point()
```

Let's add some well-written labels to the axes and legend without changing the input object. For this, we add the `labs()` function and provide a character string for each label we want to change (here `x`, `y`, and `color`):

```{r axis-labs}
ggplot(bikes, aes(y = count, x = temp, color = day_night)) + 
  geom_point() +
  labs(x = "Temperature", 
       y = "Counts of bikes shared per time period",
       color = "Period of the day")
```

You can also add each axis title via `xlab()` and `ylab()` to specify each label individually. However, I recommend to use the `labs()` as (a) it allows you to add more labels to your plot and (b) collects all labels in one place.

By adding `\n` to a string you force a line break. Usually you can also specify symbols by simply adding the symbol itself (here "°"). In case you want to add more complex expressions you can use the `expression()` function. Here we use it in combination with `paste()` to concatenate multiple strings. With this approach we can for example add superscripts:

```{r axis-labs-expression}
ggplot(bikes, aes(y = count, x = temp, color = day_night)) + 
  geom_point() +
  labs(x = expression(paste("Temperature (°C)"^"(precisely the average air temperature)")),
       y = "Counts of bikes shared per time period",
       color = "Period of\nthe day")
```
