```{r 06setupSectionTipsWorkflow, echo=FALSE, include=FALSE, freeze=FALSE}
source(here::here("src", "prerequisites.R"))
```

# Tips to Improve Your ggplot Workflow

When designing graphics with a code-based software, changes in the code need to be rendered to a graphical output to see the actual changes. The process of saving the graphic successfully in the desired format and being able to quickly inspect the adjustments are often a main source of frustration. The first tips thus cover how-to's on saving your ggplot output in the correct dimension, improving the iterative design process, and embedding non-default typefaces. 

A benefit of a code-based graphics software is the ability to automate steps to avoid redundant code and allow for consistency. By adjusting the default settings, reusing code or even writing your own helper functions and themes rather than typing the same code again and again enables one to work more efficiently while preventing mistakes and inconsistencies.


## Save ggplot Output with the "Correct" Dimensions

One of the most common question I hear is "how do I save the plot in a way that it looks as in the RStudio Plots pane?". As the Plots pane in RStudio is responsive, it will redraw the plot with new dimensions every time you resize the pane. Consequently, the relative size of geometries and theme elements changes as well. Also, the dimensions of the plot shown are not obvious so you may end up in a loop of guessing the right measurements and adjusting the sizes, margins, widths of your components. 

For sure, the iteration of adjusting sizes and position is one of the most annoying and elaborate tasks when designing graphics with **ggplot2**. And most ggplot users will painfully remember the moment when a plot saved to disk looks absolutely nothing as it did in the Plots pane. So how can we avoid this time-consuming task?

***Don't rely on the preview and decide on the dimensions before polishing:*** Once you are moving from exploring the data and potential visualizations to creating your final polished graphic, save the plot to disk in the desired width and height. By default, the dimensions are in inches.

```{r 06ggsave, eval=FALSE}
ggsave("plot.png", width = 12, height = 8)
ggsave("plot.png", width = 30, height = 20, unit = "cm")
```

From now on, stick to this dimension and adjust the components after inspecting the saved graphic. Of course, you can adjust the dimensions afterwards. Slight changes will have no big impact on the relative sizes; otherwise, you may have to resize some of your components.

If you are saving raster files such as .png and .jpeg, make sure to set the desired resolution by setting `dpi`. For web applications, images usually should be of low resolution to avoid long loading times (72 dpi is the general recommendation) while graphics used for print should be of high resolution of 300 dpi or more. The default in `ggsave()` is 300 dpi.


## Display ggplot Output with the "Correct" Dimensions

The iterative process of adjusting your graphic, saving it to disk, navigating to the folder, and looking at the plot is a tedious one. To avoid the additional workload, you can optimize your workflow by previewing your graphic in the correct dimensions without leaving RStudio.

***Work in Quarto (.qmd) or Rmarkdown (.rmd) files to inspect your plots in-line:*** In qmd and rmd formats, code is run inside individual chunks for which you can define figure dimensions by setting `fig.width` or `fig.height` as chunk options [^06-tips-1].

````{verbatim 06chunkSettings}
```{r, fig.width=12, fig.height=8}
```
````

[^06-tips-1]: Note that the new YAML-styled `#| fig-width: 12` does not work and the [width and height need to be specified in the **knitr** chunk options](https://github.com/rstudio/rstudio/issues/11708).

When a chunk is executed, the in-line plot output will respect these settings. When saving the file with `ggsave()`, make sure to use the same width and height. By default, the units in both chunk settings and `ggsave()` are in inches.  
    
**Use the camcorder package to automatically save and display your graphics:** The [**camcorder** package](https://thebioengineer.github.io/camcorder/) [@camcorder] is an R package that automatically saves graphics generated with **ggplot2** with custom settings. The package allows you to inspect the ggplot output directly in RStudio by displaying the saved graphic in the Viewer pane. As each graphic is saved automatically, running `ggsave()` after every adjustment becomes obsolote---another time-saver!

You initiate a recoding by calling `gg_record()` with the desired path to save the plots to and image settings as you specify them in `ggsave()`:

```{r 06camcorder, eval=FALSE}
install.packages("camcorder")

camcorder::gg_record(
  dir = "./plots", ## where to save the recording
  device = "png",  ## device to use to save images
  width = 12,      ## width of saved image
  height = 8,      ## height of saved image
  dpi = 300        ## dpi to use when saving image
)
```

If you want to change some settings later, use `gg_resize_film()`:

```{r 06camcorderResize, eval=FALSE}
camcorder::gg_resize_film(width = 12.5)
```

The original purpose of **camcorder** is the generation of a GIF that showcases every step of the design process using the saved image files. If you want to create such an animation, run `gg_playback()`:

```{r 06camcorderPlayback, eval=FALSE}
camcorder::gg_playback(name = "./plots/process.gif")
```


## Make Fonts Work

***Use the systemfonts package to find and use typefaces:*** As discussed in the previous chapter, the **systemfonts** package is likely the preferred approach to work with fonts. To search for a font file on your system use `match_font()`. You can inspect all available typefaces via `system_fonts()`.

A common issue is the rendering of custom typefaces and special characters such as mathematical symbols, unusual glyphs, and emojis. 

***If you are saving your graphics as PNG raster images use the agg_png device:*** The best-performing device to render your ggplot output successfully are the agg devices. In recent versions, this device is the default in `ggsave()` in case the **ragg** package is installed. It also features agg devices for other raster formats such as JPEG and TIFF.

```{r 06ggsaveAGG, eval=FALSE}
library(ragg)
ggsave("plot.png", device = agg_png)
```

To make fonts work in the RStudio Plots pane, select the AGG device as default for the graphics backend by navigating to `Tools > Global Options... > General > Graphics`.

***If you are saving your graphics as PDF vector graphics use the cairo_pdf device:*** I prefer to dave my files as lossless vector graphics. The device known to work best for rendering is the Cairo device which needs to be secified as `device = cairo_pdf`. Mac users need to install [XQuartz](https://www.xquartz.org/) which is needed to use the Cairo device.

```{r 06ggsaveCairo, eval=FALSE}
ggsave("plot.pdf", device = cairo_pdf)
```


## Set Your ggplot Theme Globally

Instead of applying and customizing a theme for each of your ggplots, you can set your theme once and it will be used for all future ggplot output by passing a complete theme to `theme_set()`.

```{r 06themeSet}
theme_set(theme_minimal())
```

Let's see how a plot using `theme_minimal()` looks like---note that we do not apply the theme to the plot itself.

```{r 06examplePlotThemeSet}
g <- 
  ggplot(bikes, aes(x = temp, y = count)) +
  geom_point(aes(color = day_night), alpha = .4) +
  ## style labels and colors
  scale_x_continuous(labels = scales::label_percent(scale = 1)) +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_color_manual(values = c("#FFA200", "#757bc7")) +
  labs(
    x = "Humidity", y = "Reported bike rents", color = NULL,
    title = "Reported bike rents versus average humidity (day/night)",
    caption = "Data: TfL (Transport for London), Jan 2015–Dec 2016"
  )
  
g
```

Furthermore, you can customize the new default theme supplied to `theme_set()`:

1. Set the typeface for text elements and sizes of the theme elements when calling the complete theme.
2. Adjust theme elements and components of the complete theme via `theme_update()`.

```{r 06themeUpdate}
theme_set(
  theme_minimal(base_family = "Asap SemiCondensed", base_size = 15)
)

theme_update(
  panel.grid.minor = element_blank(),
  plot.title = element_text(face = "bold"),
  plot.title.position = "plot",
  plot.caption = element_text(color = "grey40")
)
```

[WIP]

```{r 06examplePlotThemeUpdate, fig.cap="Running the same plot with updated theme defaults, globally set via `theme_update()`. This allows for a consistent style for all graphics without the need to write redundant code."}
g
```


## Automating Plots: Batch Operations

You can create the same plot for a range of subsets by iterating over a list of filter conditions. First, we need to specify a function that filters the data and plots the subset---and optionally also saves the output.

```{r, eval=FALSE}
plot_scatter <- function(cat, save = TRUE) {
  g <- 
    ggplot(data = filter(bikes, season %in% cat), 
           aes(x = temp, y = count, color = day_night)) +
    geom_point() +
    ## keep axis ranges consistent
    scale_x_continuous(limits = range(bikes$temp)) +
    scale_y_continuous(limits = range(bikes$count)) +
    scale_color_manual(values = c("#FFA200", "#757bc7")) +
    labs(title = stringr::str_to_title(cat), 
         x = "Temperature", y = "Bike shares", color = NULL)
   
  if (isTRUE(save)) {
    ggsave(filename = paste0("scatter_", cat, ".png"), 
           width = 7, height = 5)
  }
  
  return(g)
}
```

Afterwards, we can perform the function for multiple subsets by iterating over a vector of categories (here seasons) with the help of the **{purrr}** package. Alternatively, you can use `lapply()` or a for loop to iterate over your filter conditions.

```{r, eval=FALSE}
## print the plots as well
purrr::map(unique(bikes$season), ~plot_scatter(cat = .x))

## just run the code without printing
purrr::walk(unique(bikes$season), ~plot_scatter(cat = .x))
```

```{r 06workflowBatchSubsets, echo=FALSE, purl=FALSE, fig.width=14, fig.height=8.8, fig.cap='The four scatter plots of bike share versus temperature per season, produced by iterating our custom `plot_scatter()` function over the four different groups.'}
plot_scatter <- function(cat) {
  g <- 
    ggplot(data = filter(bikes, season %in% cat), 
           aes(x = temp, y = count, color = day_night)) +
    geom_point() +
    scale_x_continuous(limits = range(bikes$temp)) +
    scale_y_continuous(limits = range(bikes$count)) +
    scale_color_manual(values = c("#FFA200", "#757bc7")) +
    labs(title = stringr::str_to_title(cat), 
         x = "Temperature", y = "Bike shares", color = NULL) +
    theme(plot.margin = margin(rep(10, 5)))
  
  return(g)
}

patchwork::wrap_plots(
  purrr::map(unique(bikes$season), ~plot_scatter(cat = .x))
)
```

<p class="note">

Note that the axis limits differ depending on the range of the positional variables `temp` and `count`. If you want to ensure consistent ranges, add the following two lines to the plot inside the function:

```{r, eval=FALSE}
scale_x_continuous(limits = range(bikes$temp)) +
scale_y_continuous(limits = range(bikes$count)) 
```

</p>

You can leverage this approach to create the same chart for different data sets, to map different variables to aesthetics, and more.

Here is another, more general function to create a custom density plot for any quantitative variable of a data set of your choice. The function makes use of *nonstandard evaluation* to pass the variables and groupings as strings. The `!!sym()` notation allows to pass quoted variable names that are evaluated by **ggplot2**. `sym()` converts the string to a symbol, the so-called *bang-bang* `!!` then unquotes the symbol so that the `aes()` function can handle it as usual. 

```{r, eval=FALSE}
plot_density <- function(data, var, grp = "") {
  g <- 
    ggplot(data, aes(x = !!sym(var))) +
    geom_density(aes(fill = !!sym(grp)), position = "identity",
                 color = "grey30", alpha = .3) +
    coord_cartesian(expand = FALSE, clip = "off") +
    scale_y_continuous(labels = scales::label_number()) +
    scale_fill_brewer(palette = "Dark2", name = NULL) +
    theme_minimal(base_family = "Asap Condensed") +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank(),
          legend.position = "top")
    
  return(g)
}

## apply to the `mpg` data from ggplot2
purrr::map(c("displ", "hwy", "cty"), 
           ~plot_density(data = mpg, var = .x))

## apply to the `bikes` data with additional grouping
purrr::map(c("count", "temp", "humidity"), 
           ~plot_density(data = bikes, var = .x, grp = "day_night"))
```

```{r 06workflowBatchDataVar, echo=FALSE, purl=FALSE, fig.height=5, fig.cap='Two example use cases of the custom `plot_density()` function applied to three variables of the `mpg` data (upper row) and the `bikes` data (lower row), respectively. For the latter, `day_night` was passed as the grouping variable.'}
plot_density <- function(data, var, grp = "") {
  g <- 
    ggplot(data, aes(x = !!sym(var))) +
    geom_density(aes(fill = !!sym(grp)), position = "identity",
                 color = "grey30", alpha = .3) +
    coord_cartesian(expand = FALSE, clip = "off") +
    scale_y_continuous(labels = scales::label_number()) +
    scale_fill_brewer(palette = "Dark2", name = NULL) +
    theme_minimal(base_family = "Asap Condensed") +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank(),
          legend.position = "top",
          plot.margin = margin(rep(10, 5)))
    
  return(g)
}

upr <- wrap_plots(
  purrr::map(c("displ", "hwy", "cty"), 
             ~plot_density(data = mpg, var = .x),
  nrow = 1
))
           
lwr <- wrap_plots(
  purrr::map(c("count", "temp", "humidity"), 
             ~plot_density(data = bikes, var = .x, grp = "day_night"),
  nrow = 1
))

upr / plot_spacer() / lwr + plot_layout(heights = c(1, .03, 1))
```


