```{r setup-section-axes, echo=FALSE, include=FALSE}
library(tidyverse)
bikes <- readr::read_csv("https://www.cedricscherer.com/data/london-bikes.csv", col_types = "Dcfffilllddddfc")
```

# Working with Axes {#axes}

##### Increase Space between Axis and Axis Titles

`theme()` is an essential command to modify particular theme elements (texts and titles, boxes, symbols, backgrounds, ...). We are going to use them a lot! For now, we are going to modify text elements. We can change the properties of all or particular text elements (here axis titles) by overwriting the default `element_text()` within the `theme()` call:

```{r labs-move-away-vjust}
ggplot(bikes, aes(x = date, y = count)) +
  geom_point(color = "firebrick") +
  labs(x = "Date of the year", 
       y = "Counts of bikes shared per time period") +
  theme(axis.title.x = element_text(vjust = 0, size = 15),
        axis.title.y = element_text(vjust = 2, size = 15))
```

`vjust` refers to the vertical alignment, which usually ranges between 0 and 1 but you can also specify values outside that range. Note that even though we move the axis title on the y axis horizontally, we need to specify `vjust` (which is correct form the label's perspective). You can also change the distance by specifying the margin of both text elements:

```{r labs-move-away-margin}
ggplot(bikes, aes(x = date, y = count)) +
  geom_point(color = "firebrick") +
  labs(x = "Date of the year", 
       y = "Counts of bikes shared per time period") +
  theme(axis.title.x = element_text(margin = margin(t = 10), size = 15),
        axis.title.y = element_text(margin = margin(r = 10), size = 15))
```

The labels `t` and `r` within the `margin()` object refer to *top* and *right*, respectively. You can also specify the four margins as `margin(t, r, b, l)`. Note that we now have to change the right margin to modify the space on the y axis, not the bottom margin.

💡 **A good way to remember the order of the margin sides is "*t*-*r*-ou-*b*-*l*-e".**


##### Change Aesthetics of Axis Titles

Again, we use the `theme()` function and modify the element `axis.title` and/or the subordinated elements `axis.title.x` and `axis.title.y`. Within the `element_text()` we can for example overwrite the defaults for `size`, `color`, and `face`:

```{r labs-color-axes-1}
ggplot(bikes, aes(x = date, y = count)) +
  geom_point(color = "firebrick") +
  labs(x = "Date of the year", 
       y = "Counts of bikes shared per time period") +
  theme(axis.title = element_text(size = 15, color = "firebrick",
                                  face = "italic"))
```

The `face` argument can be used to make the font `bold` or `italic` or even `bold.italic`.

```{r labs-color-axes-2}
ggplot(bikes, aes(x = date, y = count)) +
  geom_point(color = "firebrick") +
  labs(x = "Date of the year", 
       y = "Counts of bikes shared per time period") +
  theme(axis.title.x = element_text(color = "sienna", size = 15),
        axis.title.y = element_text(color = "orangered", size = 15))
```

<details>
  <summary>&#128129; <i>You could also use a combination of `axis.title` and `axis.title.y`, since `axis.title.x` inherits the values from `axis.title`. Expand to see example.</i></summary>
```{r labs-color-axes-3}
ggplot(bikes, aes(x = date, y = count)) +
  geom_point(color = "firebrick") +
  labs(x = "Date of the year", 
       y = "Counts of bikes shared per time period") +
  theme(axis.title = element_text(color = "sienna", size = 15),
        axis.title.y = element_text(color = "orangered", size = 15))
```
</details>

One can modify some properties for both axis titles and other only for one or properties for each on its own:

```{r labs-color-axes-4}
ggplot(bikes, aes(x = date, y = count)) +
  geom_point(color = "firebrick") +
  labs(x = "Date of the year", 
       y = "Counts of bikes shared per time period") +
  theme(axis.title = element_text(color = "sienna", size = 15, face = "bold"),
        axis.title.y = element_text(face = "bold.italic"))
```


##### Change Aesthetics of Axis Text

Similarly, you can also change the appearance of the axis text (here *the numbers*) by using `axis.text` and/or the subordinated elements `axis.text.x` and `axis.text.y`:

```{r labs-color-axes-text}
ggplot(bikes, aes(x = date, y = count)) +
  geom_point(color = "firebrick") +
  labs(x = "Date of the year", 
       y = "Counts of bikes shared per time period") +
  theme(axis.text = element_text(color = "dodgerblue", size = 12),
        axis.text.x = element_text(face = "italic"))
```


##### Rotate Axis Text

Specifying an `angle` allows you to rotate any text elements. With `hjust` and `vjust` you can adjust the position of the text afterwards horizontally (0 = left, 1 = right) and vertically (0 = top, 1 = bottom):

```{r axis-text}
ggplot(bikes, aes(x = date, y = count)) +
  geom_point(color = "firebrick") +
  labs(x = "Date of the year", 
       y = "Counts of bikes shared per time period") +
  theme(axis.text.x = element_text(angle = 50, vjust = 1, hjust = 1, size = 12))
```


##### Remove Axis Text & Ticks

There may be rarely a reason to do so—but this is how it works:

```{r axis-no-labs}
ggplot(bikes, aes(x = date, y = count)) +
  geom_point(color = "firebrick") +
  labs(x = "Date of the year", 
       y = "Counts of bikes shared per time period") +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank())
```

I introduced three theme elements—text, lines, and rectangles—but actually there is one more: `element_blank()` which removes the element (and thus is not considered an official element).

💡 **If you want to get rid of a theme element, the element is always `element_blank()`.**


##### Remove Axis Titles

We could again use `theme_blank()` but it is way simpler to just remove the label in the `labs()` (or `xlab()`) call:

```{r axis-no-title}
ggplot(bikes, aes(x = date, y = count)) +
  geom_point(color = "firebrick") +
  labs(x = NULL, y = "")
```

💡 **Note that `NULL` removes the element (similarly to `element_blank()`) while empty quotes `""` will keep the spacing for the axis title and simply print nothing.**


##### Limit Axis Range

Sometimes you want to <s>zoom into</s> take a closer look at some range of your data. You can do this without subsetting your data:

```{r axis-limit}
ggplot(bikes, aes(x = date, y = count)) +
  geom_point(color = "firebrick") +
  labs(x = "Date of the year", 
       y = "Counts of bikes shared per time period") +
  ylim(c(0, 50))
```

Alternatively you can use `scale_y_continuous(limits = c(0, 50))` or `coord_cartesian(ylim = c(0, 50))`. The former removes all data points outside the range while the second adjusts the visible area and is similar to `ylim(c(0, 50))`. You may wonder: *So in the end both result in the same.* But not really, there is an important difference—compare the two following plots:

```{r axis-limit-comp, echo=FALSE, purl=FALSE}
p1 <- ggplot(bikes, aes(x = date, y = count)) +
  geom_point(color = "firebrick") +
  labs(x = "Date of the year", 
       y = "Counts of bikes shared per time period") +
  scale_y_continuous(limits = c(0, 5000)) +
  ggtitle("scale_y_continuous(limits) or ylim()")

p2 <- ggplot(bikes, aes(x = date, y = count)) +
  geom_point(color = "firebrick") +
  labs(x = "Date of the year", 
       y = "Counts of bikes shared per time period") +
  coord_cartesian(ylim = c(0, 5000)) +
  ggtitle("coord_cartesian(ylim)")

library(patchwork)
p1 + p2
```

You might have spotted that on the left there is some empty buffer around your y limits while on the right points are plotted right up to the border and even beyond. This perfectly illustrates the subsetting (left) versus the zooming (right). To show why this is important let's have a look at a different chart type, a box plot:

```{r axis-limit-comp-box, echo=FALSE, purl=FALSE}
p1 <- ggplot(bikes, aes(x = factor(year), y = count)) +
  geom_boxplot() +
  labs(x = "Date of the year", 
       y = "Counts of bikes shared per time period") +
  scale_y_continuous(limits = c(0, 5000)) +
  ggtitle("scale_y_continuous(limits) or ylim()")

p2 <- ggplot(bikes, aes(x = factor(year), y = count)) +
  geom_boxplot() +
  labs(x = "Date of the year", 
       y = "Counts of bikes shared per time period") +
  coord_cartesian(ylim = c(0, 5000)) +
  ggtitle("coord_cartesian(ylim)")

p1 + p2
```

Um. Because `scale_x|y_continuous()` subsets the data first, we get completely different (and wrong, at least if in the case this was not your aim) estimates for the box plots! I hope you don't have to go back to your old scripts now and check if you *maybe* have manipulated your data while plotting and did report wrong summary stats in your report, paper or thesis...


##### Force Plot to Start at Origin

Related to that, you can force R to plot the graph starting at the origin:

```{r origin}
library(tidyverse)

ggplot(bikes, aes(x = wind_speed, y = count)) +
  geom_point(color = "darkcyan") +
  labs(x = "Wind speed",
       y = "Counts of bikes shared per time period") +
  expand_limits(x = 0, y = 0)
```

<details>
  <summary>&#128129; <i>Using `coord_cartesian(xlim = c(0, NA), ylim = c(0, NA))` will lead to the same result. Expand to see example.</i></summary>
```{r origin-coord}
library(tidyverse)

ggplot(bikes, aes(x = wind_speed, y = count)) +
  geom_point(color = "darkcyan") +
  labs(x = "Wind speed",
       y = "Counts of bikes shared per time period") +
  coord_cartesian(xlim = c(0, NA), ylim = c(0, NA))
```
</details>

But we can also force it to *literally* start at the origin!

```{r origin-force}
ggplot(bikes, aes(x = wind_speed, y = count)) +
  geom_point(color = "darkcyan") +
  labs(x = "Wind speed",
       y = "Counts of bikes shared per time period") +
  expand_limits(x = 0, y = 0) +
  coord_cartesian(expand = FALSE, clip = "off")
```

💡 **The argument `clip = "off"` in any coordinate system, always starting with `coord_*`, allows to draw outside of the panel area.**

Here, I call it to make sure that the tick marks at `c(0, 0)` are not cut. See the [Twitter thread by Claus Wilke](https://twitter.com/clauswilke/status/991542952802619392?lang=en) for more details.


##### Axes with Same Scaling

For demonstrating purposes, let's plot temperature against temperature with some random noise. The `coord_equal()` is a coordinate system with a specified ratio representing the number of units on the y-axis equivalent to one unit on the x-axis. The default, `ratio = 1`, ensures that one unit on the x-axis is the same length as one unit on the y-axis:

```{r axes-equal}
ggplot(bikes, aes(x = temp, y = temp_feel)) +
  geom_point(color = "sienna") +
  labs(x = "Air temperature (°F)", y = "Feeled temperature (°F)") +
  coord_fixed()
```

Ratios higher than one make units on the y axis longer than units on the x-axis, and vice versa:

```{r axes-fixed-2, fig.height=3}
ggplot(bikes, aes(x = temp, y = temp_feel)) +
  geom_point(color = "sienna") +
  labs(x = "Air temperature (°F)", y = "Feeled temperature (°F)") +
  coord_fixed(ratio = 1/5)
```


##### Use a Function to Alter Labels

Sometimes it is handy to alter your labels a little, perhaps adding units or percent signs without adding them to your data. You can use a function in this case:

```{r labs-alt}
ggplot(bikes, aes(x = date, y = temp)) +
  geom_point() +
  labs(x = "Date of the year", y = NULL) +
  scale_y_continuous(label = function(x) {return(paste(x, "Degrees Celsius"))})
```

+ work with scale_*_date()
