```{r 05setupSectionGeoms, echo=FALSE, include=FALSE, freeze=FALSE}
source(here::here("src", "prerequisites.R")) 
```

# (PART) How To Work with Components {-}

# Working with Layers {#working-with-layers}

Internally all layers are created by the `layer()` function. A layer is by definition a "combination of data, stat and geom with a potential position adjustment" (`layer` R help). 

```{r 07layerScatter, fig.cap="A simple scatter plot of bike counts versus average temperature created with the `layer()` function."}
ggplot() +
  layer(
    data = bikes, 
    mapping = aes(x = temp, y = count),
    geom = "point", 
    stat = "identity", 
    position = "identity"
  )
```

The data and aesthetic mappings are often specified in the initial `ggplot()` call. These specifications are then passed to all layers. 

```{r, eval=FALSE}
ggplot(bikes, aes(x = temp, y = count)) +
  layer(geom = "point", stat = "identity", position = "identity")
```


## Predefined Layers

To simplify your life, **ggplot2** features a long list of predefined layers with sensible defaults for `geom`, `stat`, and `position`. Focusing on either the geometrical representation or the statistical transformation, these functions start with `geom_*()` or `stat_*()`, respectively.  

For all geometries and statistical transformation there are respective `geom_*()` and `stat_*()` functions. This means, you can usually go both routes to create the same chart. The following codes produce the same output as the `layer()` function above, a scatter plot of counts versus temperature:

```{r, eval=FALSE}
ggplot(bikes, aes(x = temp, y = count)) +
  geom_point() ## with `stat = "identity", position = "identity"`

ggplot(bikes, aes(x = temp, y = count)) +
  stat_identity() ## with `geom = "point", position = "identity"`
```

Similarly, `geom_bar()` or `stat_count()` both produce *bar* charts with the height encoding the *count* per variable. For both, the default positional adjustment is `"stack"`. Mapping a variable to the `fill` aesthetic thus creates stacked bar charts:

```{r, eval=FALSE}
ggplot(bikes, aes(x = weather_type, fill = year)) + 
  geom_bar() ## with `stat = "count"`

ggplot(bikes, aes(x = weather_type, fill = year)) + 
  stat_count() ## with `geom = "bar"`
```

Maybe you have wondered at some point, why you can use `geom_smooth()` and `stat_smooth()` interchangeably to create a conditional smoothing? This is because both call the same underlying layer function with the default arguments `geom = "smooth"` and `stat = "smooth"`, respectively. 

```{r, eval=FALSE}
ggplot(bikes, aes(x = temp, y = humidity)) + 
  geom_smooth() ## with `stat = "smooth"`

ggplot(bikes, aes(x = temp, y = humidity)) + 
  stat_smooth() ## with `geom = "smooth"`
```

```{r 07layerPredefinedDefault, echo=FALSE, fig.cap="Left: A stacked bar chart of the number of observations per weather type and time of the day, created with the predefined layers `geom_bar()` or `stat_count()`. Right: A conditional smoothing of humidity and temperature created with the predefined layers `geom_smooth()` or `stat_smooth()`."}
s1 <- 
  ggplot(bikes, aes(x = weather_type, fill = day_night)) + 
  stat_count() +
  scale_fill_manual(values = prismatic::clr_darken(colors[2:3], .2), name = NULL) +
  labs(title = "geom_bar()\nstat_count()") +
  theme(plot.title = element_text(family = "Spline Sans Mono", size = rel(1.05), lineheight = 1.1), 
        panel.grid.major.x = element_blank(),
        legend.position = "bottom")

s2 <- 
  ggplot(bikes, aes(x = temp, y = humidity)) + 
  geom_smooth(color = "black") +
  labs(title = "geom_smooth()\nstat_smooth()") +
  theme(plot.title = element_text(family = "Spline Sans Mono", size = rel(1.05), lineheight = 1.1))

s1 + plot_spacer() + s2 + plot_layout(widths = c(1, .03, .8))
```

## Changing Layer Defaults

We can overwrite the default argument inputs to change the behavior of the layer. 

While the stacked bars (Fig. \@ref(fig:07layerPredefinedDefault)A) put emphasis on the total counts per weather type, a grouped bar chart simplifies comparison between groups. We can change the default behavior of `geom_bar()` or `stat_count()` by setting the `position` to `dodge` to place the bars next to each other:

```{r, eval=FALSE}
ggplot(bikes, aes(x = weather_type, fill = day_night)) +
  geom_bar(position = "dodge")
```

We can also overwrite the default `stat` in geometrical layers or `geom` in statistical layers. For example, changing the `geom` in `stat_smooth()` to `pointrange` turns the smoothed line and ribbon into points with vertical lines representing the confidence intervals:

```{r, eval=FALSE}
ggplot(bikes, aes(x = wind_speed, y = humidity)) +
  stat_smooth(geom = "pointrange")
```

Again, we could rewrite the code by using the respective geometrical layer `geom_pointrange()` and setting the default statistical transformation `"identity"` to `"smooth"`.

```{r 07layerPredefinedCustom, echo=FALSE, fig.cap='Left: The bar chart of weather types per time of the day as dodged bars by setting the `position` to `"dodge"`, overwriting the default `"stack"`. Right: The conditional smoothing as so-called pointranges displaying the predicted mean as points and the confidence intervals as vertical lines by overwriting the default `geom` in `stat_smooth()` or using the respective geometrical layer with `stat = "smooth"`.'}
s3 <- ggplot(bikes, aes(x = weather_type, fill = day_night)) +
  geom_bar(position = position_dodge(preserve = "single")) +
  scale_fill_manual(values = prismatic::clr_darken(colors[2:3], .2), name = NULL) +
  labs(title = 'geom_bar(position = "dodge")\nstat_count(position = "dodge")') +
  theme(plot.title = element_text(family = "Spline Sans Mono", size = rel(1.05), lineheight = 1.1), 
        panel.grid.major.x = element_blank(),
        legend.position = "bottom")

s4 <- ggplot(bikes, aes(x = wind_speed, y = humidity)) +
  stat_smooth(geom = "pointrange", size = .1)  +
  labs(title = 'stat_smooth(geom = "pointrange")\nstat_pointrange(stat = "smooth")') +
  theme(plot.title = element_text(family = "Spline Sans Mono", size = rel(1.05), lineheight = 1.1))

s3 + plot_spacer() + s4 + plot_layout(widths = c(1, .03, .8))
```


## Positional Aesthetics

Many geometrical shapes work with one or two positional arguments, namely `x` and `y` to represent your data. Depending on the type of the variables, the layer may behave differently or return an error if the variables do not meet the required type. 

For example, **ggplot2** will calculate a box plot for each category if the type is `character`, `logical`, or `factor` but only a single box plot otherwise:

```{r, eval=FALSE}
## creates four boxplots as `season` is a factor
ggplot(bikes, aes(x = season, y = count)) + geom_boxplot()

## creates two boxplots as `is_weekend` is logical
ggplot(bikes, aes(x = is_weekend, y = count)) + geom_boxplot()

## creates a single boxplot as `temp` is numeric
ggplot(bikes, aes(x = temp, y = count)) + geom_boxplot()
```

```{r, echo=FALSE}
b1 <- ggplot(bikes, aes(x = season, y = count)) + geom_boxplot()
b2 <- ggplot(bikes, aes(x = is_weekend, y = count)) + geom_boxplot()
b3 <- ggplot(bikes, aes(x = temp, y = count)) + geom_boxplot()

b1 + plot_spacer() + b2 + plot_spacer() + b3 + plot_layout(widths = c(1, .05, .65, .05, .5))
```

Histograms show the distribution of numeric variables as bars grouped into equal numerical ranges, so-called bins. The corresponding `geom_histogram()` creates *binned bars* (read: `stat = "bin", geom = "bar"`) for quantitative variables but fails when passing qualitative variables as calculation of bins across categories is not meaningful. In case of a discrete x aesthetic, `geom_bar()` with the default `stat = "count"` creates the desired bar chart showing counts per category. 

```{r, eval=FALSE}
## works as `temp` is numeric
ggplot(bikes, aes(x = temp)) + geom_histogram()

## fails as `is_weekend` is categorical
ggplot(bikes, aes(x = is_weekend)) + geom_histogram()
# Error: `stat_bin()` requires a continuous x aesthetic

## create bars with discrete x aesthetics
ggplot(bikes, aes(x = is_weekend)) + geom_bar()
```

```{r, echo=FALSE}
h1 <- ggplot(bikes, aes(x = temp)) + geom_histogram()
h2 <- ggplot(bikes, aes(x = is_weekend)) + geom_bar()

h1 + plot_spacer() + h2 + plot_layout(widths = c(1, .03, .65))
```

Other geometries need more than two positional aesthetics. `geom_linerange()` draws horizontal or vertical lines based on `xmin` and `xmax` or `ymin` and `ymax`, respectively. Let's load some stock data and show the change of opening versus closing price over time:

```{r}
coins <- readr::read_rds("./data/crypto_cleaned.rds") %>% 
  dplyr::filter(year == 2018, month %in% 5:6, currency == "litecoin")

ggplot(coins, aes(x = date, ymin = open, ymax = close)) +
  geom_linerange()
```

That's great but I'd like to indicate the direction of change in daily prices. For that we can use `geom_segment()` which allows to add arrow heads to straight lines. The segments are not restricted to vertical or horizontal and thus take four positional arguments: `x` and `xend` as well as `y` and `yend`. For our use case, we map the same variable, namely `date`, to both x coordinates.

```{r}
ggplot(coins, aes(x = date, xend = date, 
                  y = open, yend = close)) +
  geom_segment(
    arrow = arrow(type = "closed",length = unit(.4, "lines"))
  )
```

The geom `pointrange` (that we have also used in the smoothing example) needs four positional aesthetics as well: in addition to the `x` and `y` positions, we have to specify the range as well as either `xmin` and `xmax` or `ymin` and `ymax`.

```{r}
library(dplyr)
bikes_season <- 
  bikes %>% 
  group_by(season) %>% 
  summarize(avg = mean(count), sd = sd(count))

ggplot(bikes_season, 
       aes(x = season, y = avg, ymin = avg - sd, ymax = avg + sd)) +
  geom_pointrange()
```

## Statistical Transformations

You can also use `stat_summary()` and let **ggplot2** do the transformation of the raw data. By default, `stat_summary()` uses a `pointrange` geometry with the three positional aesthetics representing mean ± standard error. 

```{r, eval=FALSE}
ggplot(bikes, aes(x = season, y = count)) +
  stat_summary()
```
 
Passing a set of functions to the arguments `fun`, `fun.min` and `fun.max` we can modify the range:

```{r, eval=FALSE}
ggplot(bikes, aes(x = season, y = count)) +
  stat_summary(
    fun = "mean",
    fun.min = function(y) mean(y) - sd(y), 
    fun.max = function(y) mean(y) + sd(y)
  )
```

```{r, echo=FALSE}
e1 <- ggplot(bikes, aes(x = season, y = count)) +
  stat_summary() +
  labs(title = "mean ± standard error")

e2 <- ggplot(bikes, aes(x = season, y = count)) +
  stat_summary(
    fun = "mean",
    fun.min = function(y) mean(y) - sd(y), 
    fun.max = function(y) mean(y) + sd(y)
  ) +
  labs(title = "mean ± standard deviation")

e1 + plot_spacer() + e2 + plot_layout(widths = c(1, .1, 1))
```

Alternatively, a function that returns the required aesthetics can be specified as input for `fun.data`. As the function `mean_sdl()` computes the mean ± 2 times the standard deviation, we have to adjust the function argument `mult` as well. Either we pass it to the `fun.args` argument or specify it as function argument directly.

```{r, eval=FALSE}
ggplot(bikes, aes(x = season, y = count)) +
  stat_summary(
    fun.data = "mean_sdl", 
    fun.args = list(mult = 1)
  )

ggplot(bikes, aes(x = season, y = count)) +
  stat_summary(
    fun.data = function(y) mean_sdl(y, mult = 1)
  )
```

This seems to hard to remember? The nice thing is that all  solutions are valid, just pick your favorite (and forget about the rest as long as there is no particular reason to use it).

Similarly, we can also overwrite the default geometry, for example to display the median count for each season by adding a red triangle:

```{r}
ggplot(bikes, aes(x = season, y = count)) +
  stat_summary(
    fun.data = function(y) mean_sdl(y, mult = 1)
  ) +
  stat_summary(
    geom = "point",
    fun = "median",
    color = "#FF5477",
    shape = 17,
    size = 4
  )
```

The `stat_*()` functions and especially `stat_summary` are incredible powerful and often a neat way to create all kind of chart types. Let's use a combination of summaries to draw a dumbbell chart of temperature ranges across months.

```{r}
ggplot(bikes, aes(y = month, x = temp)) + 
  stat_summary(
    geom = "linerange",
    fun.min = "min",
    fun.max = "max"
  ) +
  stat_summary(
    geom = "point",
    fun = "range",
    size = 3
  )
```

We can also use `stat_summary()` to add the actual minimum and maximum values to our plot. The geom `text` in combination with the `label` aesthetic adds data-related labels. 

```{r, eval=FALSE}
ggplot(bikes, aes(y = forcats::fct_rev(month), x = temp)) + 
  stat_summary(
    geom = "linerange",
    fun.min = "min",
    fun.max = "max"
  ) +
  stat_summary(
    geom = "point",
    fun = "range",
    size = 3
  ) +
  stat_summary(
    geom = "text",
    fun = "range",
    aes(label = after_stat(sprintf("%1.1f", x))),
    vjust = 1.7
  )
```

```{r, echo=FALSE}
ggplot(bikes, aes(y = forcats::fct_rev(month), x = temp)) + 
  stat_summary(
    geom = "linerange",
    fun.min = "min",
    fun.max = "max"
  ) +
  stat_summary(
    geom = "point",
    fun = "range",
    size = 3
  ) +
  stat_summary(
    geom = "text",
    fun = "range",
    aes(label = after_stat(sprintf("%1.1f", x))),
    family = "Asap Condensed",
    vjust = 1.7
  )
```



```{r, include=FALSE}
ggplot(bikes, aes(x = temp, y = count, group = cut_interval(temp, 4))) + 
  geom_boxplot()

ggplot(bikes, aes(x = temp, y = count, group = cut_number(temp, 4))) + 
  geom_boxplot()

ggplot(bikes, aes(x = cut_interval(temp, 4), y = count)) + 
  geom_boxplot(varwidth = TRUE)

ggplot(bikes, aes(x = cut_number(temp, 4), y = count)) + 
  geom_boxplot(varwidth = TRUE)
```



[WIP]



