```{r 05setupSectionGeoms, echo=FALSE, include=FALSE}
source(here::here("src", "prerequisites.R"))
```

# (PART) How To Work with Components {-}

# Working with Layers

Internally all layers are created by the `layer()` function. A layer is by definition a "combination of data, stat and geom with a potential position adjustment" (`layer` R help). 

The long list of featured `geom_*()` and `stat_*()` functions in **{ggplot2}** is thus just a set of predefined layers with default `geom`, `stat`, `position` arguments that use the data passed in the initial `ggplot()` call. The data can also be specified for each layer separately by passing it inside the respective geom or stat.

## Geometrical Shapes

Many geometrical shapes only need two positional arguments, namely `x` and `y` as a single coordinate is sufficient to represent your data. A combination of x and y values determine the position of points, connected lines, or the height of bars.

```{r}
ggplot(bikes, aes(x = date, y = count)) +
  geom_line() ## a timeseries of bike shares per date

ggplot(bikes, aes(x = date, y = count)) +
  geom_col() ## a bar chart with one bar per date
```

[WIP]

* most geoms use two positional aesthetics
* showcase a few of those
* some geoms take only one or more than 2
* showcase of some of those


## Statistical Transformations

* stats are just a different perspective
* illustrate with usual examples
* powerful transformations: count + summary


## Positional Adjustments

* 


